---
title: "Delay Discounting: Mitchell Lab"
author: "Montana Kay Lara"
date: "2023-09-20"
output: html_document
---

Update: September 27, 2023

The following markdown is used for delay discounting analysis. From the adjusting amount procedure done in HS rats, raw data should include indifference points for different time delays 'T'. Each animal is identified by 'rfid' codes. For each delay curve for each animal, this script will calculate the AUC and fit both hyperbolic and exponential functions. Additionally, each function is fit both with and wihtout bias. The resulting data table will have calculated bias, AUC, hyperbolic k parameter, exponential k parameter, and r^2 values for the models. These values are compared at the end, and these trait measures for delay discounting are used in GWAS. 

I. Setup
II. Import raw dataset and cleanup
III. Calculate bias (b) 
IV. AUC
V. Hyperbolic with bias
VI. Hyperbolic without bias
VII. Exponential with bias
VIII. Expontneital without bias
IX. Checking values


## I. Setup
This chunk imports all the necessary packages to run the rest of the code. (If you do not have the packages installed, use the function 'install.packages()')

```{r setup, results='hide', message=FALSE}
all.packages <- c("tidyverse", 
                  "ggplot2", 
                  "here", 
                  "readxl",
                  "conflicted",
                  "readr",
                  "ggdist",
                  "ggbeeswarm",
                  "ggforce", 
                  "ggpubr",
                  "readr",
                  "minpack.lm", 
                  "AICcmodavg", 
                  "e1071")
for(i in 1:length(all.packages)){library(all.packages[i], character.only = TRUE)} 
conflict_prefer("filter", "dplyr")
```


## II. Import dataset for dd gwas

The import requires a spreadsheet (either excel or csv), with: 
* rows = animals
* columns = traits

```{r import}
#processed_data_ready was taken from the Palmer lab csv used in GWAS, which has calculated values from the Mitchell lab
dd_processeddata <- read_csv("data/processed_data_ready.csv", col_types = cols(rfid = col_character()))
#dd_rawdata has just the raw values for the delays and the animal information
dd_rawdata <- read_csv("data/dd_rawdata.csv", col_types = cols(rfid = col_character()))
```

```{r comparisonNEW}
#add spreadsheet from suzanne that had NEW (correct) values
dd_newdata <- read_csv("data/parameters_forGWAS_paper.csv", col_types = cols(rfid = col_character()))

#had to create "full_rfid" to be able to match values based on rfid in all the dataframes
dd_newdata$full_rfid <- dd_newdata$rfid
for (i in 1:nrow(dd_newdata)) {
  partial <- dd_newdata$rfid[i]
  matching_row <- sapply(dd_processeddata$rfid, function(full_rfid) grepl(partial, full_rfid))
  if (any(matching_row)) {
    dd_newdata$full_rfid[i] <- dd_processeddata$rfid[matching_row]
  }
}
dd_newdata$rfid <- dd_newdata$full_rfid
```


```{r filterNA}
#create new dataframe with only the animals who have values for indifference points (filtering out all NAs)
dd_data <- dd_rawdata %>% 
  filter_at(vars(dd_indiff_0, dd_indiff_2, dd_indiff_4, dd_indiff_8, dd_indiff_16, dd_indiff_24), all_vars(!is.na(.)))
```



## III. Calculate bias
Bias (b) is calculated as the indifference point at delay 0s divided by the fixed delay reward (150ul)

```{r bias}
dd_data$bias <- dd_data$dd_indiff_0/150
```

```{r resultsdf}
#this is just to create a new dataframe to store results in
dd_fittedmodels <- dd_data
```


## IV. Melt data
This chunk creates the dataframe "dd_pivot" which takes the delays and puts them into a single column "T" and takes each indifference point value and puts them into another column "indiff_point" corresponding to the Delay T. Once the data is structured this way, we can use the functions in the subsequent chunks to fit the models. 

```{r}
dd_pivot <- dd_data %>% 
  pivot_longer(cols = dd_indiff_0:dd_indiff_24,
               names_to = "Delay", 
               values_to = "indiff_point") %>% 
  mutate(Delay = factor(Delay, levels = c("dd_indiff_0", "dd_indiff_2", "dd_indiff_4", "dd_indiff_8", "dd_indiff_16", "dd_indiff_24"))) %>% 
  mutate(T = case_when(Delay == "dd_indiff_0" ~ 0, 
                       Delay == "dd_indiff_2" ~ 2, 
                       Delay == "dd_indiff_4" ~ 4,
                       Delay == "dd_indiff_8" ~ 8,
                       Delay == "dd_indiff_16" ~ 16,
                       Delay == "dd_indiff_24" ~ 24))
```




## IV. AUC

```{r AUC}
#this creates the values for the trapezoids from NORMALIZED T values (which is the T/24)
trap0_2 <- (2/24) - (0/24)
trap2_4 <- (4/24) - (2/24)
trap4_8 <- (8/24) - (4/24)
trap8_16 <- (16/24) - (8/24)
trap16_24 <- (24/24) - (16/24)

#traparea calculates the area of the trapezoid for each time bin
#AUC is the sum of those trapezoids
dd_AUC <- dd_data %>% 
  mutate("traparea0_2" = trap0_2 * ((dd_indiff_0/150 + dd_indiff_2/150)/2), 
         "traparea2_4" = trap2_4 * ((dd_indiff_2/150 + dd_indiff_4/150)/2), 
         "traparea4_8" = trap4_8 * ((dd_indiff_4/150 + dd_indiff_8/150)/2), 
         "traparea8_16" = trap8_16 * ((dd_indiff_8/150 + dd_indiff_16/150)/2), 
         "traparea16_24" = trap16_24 * ((dd_indiff_16/150 + dd_indiff_24/150)/2)) %>% 
  mutate("AUC" = traparea0_2 + traparea2_4 + traparea4_8 + traparea8_16 + traparea16_24) %>% 
  select(rfid, traparea0_2:AUC)


#add AUC to results dataframe
dd_fittedmodels <- dd_fittedmodels %>% 
  left_join(dd_AUC, by = 'rfid')
```






## V. Hyperbolic model with bias

First we define the hyperbolic function, including the bias variable "b"
```{r hypbias}
# Defining the hyperbolic model function with A = 150
A <- 150
hyperbolic_bias_model <- function(T, b, A, k) {
  return((b * A) / (1 + k * T))}
```

This chunk fits the hyperbolic model with bias and iterates over each animal. The results is a list of k, residual sum of squares, and r^2 values for each animal
```{r fithypbias}
# the nlsLM() function will fit the hyperbolic model and start with a k value of 0
initial_k <- 0
# initialize an empty vector where the k values, sum of squared (SS) residuals, and r^2 values will be stored
best_k_values_hypbias <- list()
residualSS_hypbias <- list()
r2_hypbias <- list()

# Iterate over each unique animal rfid to fit the hyperbolic function
unique_rfids <- unique(dd_pivot$rfid)
for (single_rfid in unique_rfids) {
  animal_data <- subset(dd_pivot, rfid == single_rfid)
  # Fit the model with 'b' fixed for the current animal
  fit <- nlsLM(indiff_point ~ hyperbolic_bias_model(T, bias, A, k), data = animal_data, start = list(k = initial_k))
  # Store the best 'k' value for the current animal
  best_k_values_hypbias[[single_rfid]] <- coef(fit)
  #calculate residuals
  residuals_hypbias <- residuals(fit)
  sum_residuals_hypbias <- sum(residuals_hypbias^2)
  r_squared <- 1 - (sum_residuals_hypbias / sum((animal_data$indiff_point - mean(animal_data$indiff_point))^2))
  # store SS residuals for the current animal
  residualSS_hypbias[[single_rfid]] <- sum_residuals_hypbias
  # store r2 values
  r2_hypbias[[single_rfid]] <- r_squared
}
```

This chunk takes the lists of values for each animal and formats them to be able to be put into a new dataframe that can be accessed for analysis 
```{r dfhypbias}
#make lists into dataframes/vectors
hyperbolic_bias_k <- as.data.frame(do.call(rbind, best_k_values_hypbias))[["k"]]
hyperbolic_bias_residualSS <- as.data.frame(do.call(rbind, residualSS_hypbias))[["V1"]]
hyperbolic_bias_r2 <- as.data.frame(do.call(rbind, r2_hypbias))[["V1"]]
#put it all together into a dataframe
hypbias_results <- data.frame(rfid = unique_rfids, hyperbolic_bias_k, hyperbolic_bias_residualSS, hyperbolic_bias_r2)

#add hyp bias to results dataframe
dd_fittedmodels <- dd_fittedmodels %>% 
  left_join(hypbias_results, by = 'rfid')
```






## VI. Hyperbolic model withOUT bias

First we define the hyperbolic function, (without the bias variable "b")
```{r hyp}
# Defining the hyperbolic model function with A = 150
A <- 150
hyperbolic_model <- function(T, A, k) {
  return((A) / (1 + k * T))}
```

This chunk fits the hyperbolic model (without bias) and iterates over each animal. The results is a list of k, residual sum of squares, and r^2 values for each animal
```{r fithyp}
# the nlsLM() function will fit the hyperbolic model and start with a k value of 0
initial_k <- 0
# initialize an empty vector where the k values, sum of squared (SS) residuals, and r^2 values will be stored
best_k_values_hyp <- list()
residualSS_hyp <- list()
r2_hyp <- list()

# Iterate over each unique animal rfid to fit the hyperbolic function
unique_rfids <- unique(dd_pivot$rfid)
for (single_rfid in unique_rfids) {
  animal_data <- subset(dd_pivot, rfid == single_rfid)
  # Fit the model for the current animal
  fit <- nlsLM(indiff_point ~ hyperbolic_model(T, A, k), data = animal_data, start = list(k = initial_k))
  # Store the best 'k' value for the current animal
  best_k_values_hyp[[single_rfid]] <- coef(fit)
  #calculate residuals
  residuals_hyp <- residuals(fit)
  sum_residuals_hyp <- sum(residuals_hyp^2)
  r_squared <- 1 - (sum_residuals_hyp / sum((animal_data$indiff_point - mean(animal_data$indiff_point))^2))
  # store SS residuals for the current animal
  residualSS_hyp[[single_rfid]] <- sum_residuals_hyp
  # store r2 values
  r2_hyp[[single_rfid]] <- r_squared
}
```

This chunk takes the lists of values for each animal and formats them to be able to be put into a new dataframe that can be accessed for analysis 
```{r dfhyp}
#make lists into dataframes/vectors
hyperbolic_k <- as.data.frame(do.call(rbind, best_k_values_hyp))[["k"]]
hyperbolic_residualSS <- as.data.frame(do.call(rbind, residualSS_hyp))[["V1"]]
hyperbolic_r2 <- as.data.frame(do.call(rbind, r2_hyp))[["V1"]]
#put it all together into a dataframe
hyp_results <- data.frame(rfid = unique_rfids, hyperbolic_k, hyperbolic_residualSS, hyperbolic_r2)

#add hyp bias to results dataframe
dd_fittedmodels <- dd_fittedmodels %>% 
  left_join(hyp_results, by = 'rfid')
```










## VII. Exponential model with bias

First we define the exponential function, including the bias variable "b"
```{r expbias}
# Defining the exponential model function with A = 150
A <- 150
exponential_bias_model <- function(T, b, A, k) {
  return(b * A * exp(-k * T))}
```

This chunk fits the exponential model with bias and iterates over each animal. The results is a list of k, residual sum of squares, and r^2 values for each animal
```{r fitexpbias}
# the nlsLM() function will fit the exponential model and start with a k value of 0
initial_k <- 0
# initialize an empty vector where the k values, sum of squared (SS) residuals, and r^2 values will be stored
best_k_values_expbias <- list()
residualSS_expbias <- list()
r2_expbias <- list()

# Iterate over each unique animal rfid to fit the exponential function
unique_rfids <- unique(dd_pivot$rfid)
for (single_rfid in unique_rfids) {
  animal_data <- subset(dd_pivot, rfid == single_rfid)
  # Fit the model with 'b' fixed for the current animal
  fit <- nlsLM(indiff_point ~ exponential_bias_model(T, bias, A, k), data = animal_data, start = list(k = initial_k))
  # Store the best 'k' value for the current animal
  best_k_values_expbias[[single_rfid]] <- coef(fit)
  #calculate residuals
  residuals_expbias <- residuals(fit)
  sum_residuals_expbias <- sum(residuals_expbias^2)
  r_squared <- 1 - (sum_residuals_expbias / sum((animal_data$indiff_point - mean(animal_data$indiff_point))^2))
  # store SS residuals for the current animal
  residualSS_expbias[[single_rfid]] <- sum_residuals_expbias
  # store r2 values
  r2_expbias[[single_rfid]] <- r_squared
}
```

This chunk takes the lists of values for each animal and formats them to be able to be put into a new dataframe that can be accessed for analysis 
```{r dfexpbias}
#make lists into dataframes/vectors
exponential_bias_k <- as.data.frame(do.call(rbind, best_k_values_expbias))[["k"]]
exponential_bias_residualSS <- as.data.frame(do.call(rbind, residualSS_expbias))[["V1"]]
exponential_bias_r2 <- as.data.frame(do.call(rbind, r2_expbias))[["V1"]]
#put it all together into a dataframe
expbias_results <- data.frame(rfid = unique_rfids, exponential_bias_k, exponential_bias_residualSS, exponential_bias_r2)

#add exp bias to results dataframe
dd_fittedmodels <- dd_fittedmodels %>% 
  left_join(expbias_results, by = 'rfid')
```





## VIII. Exponential model withOUT bias

First we define the exponential function, (without the bias variable "b")
```{r exp}
# Defining the exponential model function with A = 150
A <- 150
exponential_model <- function(T, A, k) {
  return(A * exp(-k * T))}
```

This chunk fits the exponential model (without bias) and iterates over each animal. The results is a list of k, residual sum of squares, and r^2 values for each animal
```{r fitexp}
# the nlsLM() function will fit the exponential model and start with a k value of 0
initial_k <- 0
# initialize an empty vector where the k values, sum of squared (SS) residuals, and r^2 values will be stored
best_k_values_exp <- list()
residualSS_exp <- list()
r2_exp <- list()

# Iterate over each unique animal rfid to fit the exponential function
unique_rfids <- unique(dd_pivot$rfid)
for (single_rfid in unique_rfids) {
  animal_data <- subset(dd_pivot, rfid == single_rfid)
  # Fit the model for the current animal
  fit <- nlsLM(indiff_point ~ exponential_model(T, A, k), data = animal_data, start = list(k = initial_k))
  # Store the best 'k' value for the current animal
  best_k_values_exp[[single_rfid]] <- coef(fit)
  #calculate residuals
  residuals_exp <- residuals(fit)
  sum_residuals_exp <- sum(residuals_exp^2)
  r_squared <- 1 - (sum_residuals_exp / sum((animal_data$indiff_point - mean(animal_data$indiff_point))^2))
  # store SS residuals for the current animal
  residualSS_exp[[single_rfid]] <- sum_residuals_exp
  # store r2 values
  r2_exp[[single_rfid]] <- r_squared
}
```

This chunk takes the lists of values for each animal and formats them to be able to be put into a new dataframe that can be accessed for analysis 
```{r dfexp}
#make lists into dataframes/vectors
exponential_k <- as.data.frame(do.call(rbind, best_k_values_exp))[["k"]]
exponential_residualSS <- as.data.frame(do.call(rbind, residualSS_exp))[["V1"]]
exponential_r2 <- as.data.frame(do.call(rbind, r2_exp))[["V1"]]
#put it all together into a dataframe
exp_results <- data.frame(rfid = unique_rfids, exponential_k, exponential_residualSS, exponential_r2)

#add exp bias to results dataframe
dd_fittedmodels <- dd_fittedmodels %>% 
  left_join(exp_results, by = 'rfid')
```






## IX. Checking values


```{r}
write.csv(dd_fittedmodels, here("results/dd_modelresults.csv"))
```


```{r}
models <- list(age.mod, sex.mod, consumption.mod, age.sex.mod, combination.mod, interaction.mod)

model.names <- c('age.mod', 'sex.mod', 'consumption.mod', 'age.sex.mod', 'combination.mod', 'interaction.mod')
```


```{r}
test_merged_df <- merge(dd_newdata, dd_AUC, by = "rfid", suffixes = c("_df1", "_df2"), all = TRUE)
ggplot(test_merged_df, aes(x = `ObjectiveAUC-normD`, y = AUC)) +
  geom_point() +
  stat_cor(method = "pearson")

ggplot(dd_fittedmodels, aes(x = hyperbolic_k, y = AUC)) +
  geom_point() +
  stat_cor(method = "pearson")
```

```{r}
skewness(dd_fittedmodels$dd_indiff_2)
kurtosis(dd_fittedmodels$dd_indiff_2)
hist(dd_fittedmodels$dd_indiff_2)
```
```{r}
skewness(dd_fittedmodels$AUC)
kurtosis(dd_fittedmodels$AUC)
hist(dd_fittedmodels$AUC)
```

```{r}
skewness(dd_fittedmodels$exponential_bias_k)
kurtosis(dd_fittedmodels$exponential_bias_k)
hist(dd_fittedmodels$exponential_bias_k)
```
```{r}
skewness(dd_fittedmodels$exponential_k)
kurtosis(dd_fittedmodels$exponential_k)
hist(dd_fittedmodels$exponential_k)
```

```{r}
skewness(dd_fittedmodels$hyperbolic_bias_k)
kurtosis(dd_fittedmodels$hyperbolic_bias_k)
hist(dd_fittedmodels$hyperbolic_bias_k)
```


```{r}
skewness(dd_fittedmodels$hyperbolic_k)
kurtosis(dd_fittedmodels$hyperbolic_k)
hist(dd_fittedmodels$hyperbolic_k)
```




